---
title: "Safety Trends"
author: "PSRC Safety Summit"
date: "June 29, 2023"
output: 
  ioslides_presentation:
    widescreen: true
    css: styles.css
runtime: shiny
---

```{r setup, include=FALSE}
library(tidyverse)
library(echarts4r)
library(psrcplot)
library(leaflet)
library(sf)

knitr::opts_chunk$set(echo = FALSE)
e_common(font_family = "Poppins")
data_file <- "C:/Users/chelmann/OneDrive - Puget Sound Regional Council/coding/psrc-safety-plan/convening/data/collision_data.csv"

mpo_file <- "C:/Users/chelmann/OneDrive - Puget Sound Regional Council/coding/psrc-safety-plan/convening/data/mpo_data.csv"

psrc.logo <- "data/psrc_logo.png"

# Custom svg icons for pictorial charts
fa_house <- "path://M575.8 255.5c0 18-15 32.1-32 32.1h-32l.7 160.2c0 2.7-.2 5.4-.5 8.1V472c0 22.1-17.9 40-40 40H456c-1.1 0-2.2 0-3.3-.1c-1.4 .1-2.8 .1-4.2 .1H416 392c-22.1 0-40-17.9-40-40V448 384c0-17.7-14.3-32-32-32H256c-17.7 0-32 14.3-32 32v64 24c0 22.1-17.9 40-40 40H160 128.1c-1.5 0-3-.1-4.5-.2c-1.2 .1-2.4 .2-3.6 .2H104c-22.1 0-40-17.9-40-40V360c0-.9 0-1.9 .1-2.8V287.6H32c-18 0-32-14-32-32.1c0-9 3-17 10-24L266.4 8c7-7 15-8 22-8s15 2 21 7L564.8 231.5c8 7 12 15 11 24z"

fa_person <- "path://M112 48a48 48 0 1 1 96 0 48 48 0 1 1 -96 0zm40 304V480c0 17.7-14.3 32-32 32s-32-14.3-32-32V256.9L59.4 304.5c-9.1 15.1-28.8 20-43.9 10.9s-20-28.8-10.9-43.9l58.3-97c17.4-28.9 48.6-46.6 82.3-46.6h29.7c33.7 0 64.9 17.7 82.3 46.6l58.3 97c9.1 15.1 4.2 34.8-10.9 43.9s-34.8 4.2-43.9-10.9L232 256.9V480c0 17.7-14.3 32-32 32s-32-14.3-32-32V352H152z"

fa_user <- "path://M256 288A144 144 0 1 0 256 0a144 144 0 1 0 0 288zm-94.7 32C72.2 320 0 392.2 0 481.3c0 17 13.8 30.7 30.7 30.7H481.3c17 0 30.7-13.8 30.7-30.7C512 392.2 439.8 320 350.7 320H161.3z"

collisions <- read_csv(data_file, show_col_types = FALSE) %>%
  mutate(injury_type = str_replace_all(injury_type, "Fatality", "Traffic Related Deaths")) %>%
  mutate(injury_type = str_replace_all(injury_type, "Serious Injury", "Serious Injuries")) %>%
  mutate(injury_rate_per_100k = round(injury_rate_per_100k,1)) %>%
  mutate(collision_rate_per_100k = round(collision_rate_per_100k,1))

num_cities <- collisions %>%
  filter(year >= 2018 & geography=="City") %>%
  select(name) %>%
  distinct() %>%
  pull() %>%
  length()

metros <- read_csv(mpo_file, show_col_types = FALSE) 

tracts <- st_read("https://services6.arcgis.com/GWxg6t7KXELn1thE/arcgis/rest/services/Census_Tracts_2010/FeatureServer/0/query?where=0=0&outFields=*&f=pgeojson") %>% select(name="geoid10")

tract_collisions <- collisions %>% 
  filter(geography == "Census Tract 2010" & year >= 2018) %>%
  group_by(name) %>%
  summarise(injuries = sum(injuries)) %>%
  as_tibble()

tracts <- left_join(tracts, tract_collisions, by=c("name")) %>%
  mutate(injuries = replace_na(injuries, 0))

create_tract_map <-function(tract_layer, d_title) {
  
  title <- tags$div(HTML(d_title))
  
  # Determine Bins
  rng <- range(tract_layer$injuries)
  max_bin <- max(abs(rng))
  round_to <- 10^floor(log10(max_bin))
  max_bin <- ceiling(max_bin/round_to)*round_to
  breaks <- (max_bin*c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.8, 1))
  bins <- c(0, breaks)
  
  pal <- colorBin(psrc_colors$purples_inc, domain = tract_layer$injuries, bins = bins)
  labels <- paste0("<b>",paste0("Serious or Fatal Injuries: "), "</b>", tract_layer$injuries) %>% lapply(htmltools::HTML)
  
  working_map <- leaflet(data = tract_layer, options = leafletOptions(zoomControl=FALSE)) %>% 
    addProviderTiles(providers$CartoDB.Positron) %>%
    addLayersControl(baseGroups = c("Base Map"),
                     overlayGroups = c("Census Tracts"),
                     options = layersControlOptions(collapsed = TRUE)) %>%
    addPolygons(fillColor = pal(tract_layer$injuries),
                weight = 1.0,
                opacity = 1,
                color = "white",
                dashArray = "3",
                fillOpacity = 0.7,
                highlight = highlightOptions(
                  weight =5,
                  color = "76787A",
                  dashArray ="",
                  fillOpacity = 0.7,
                  bringToFront = TRUE),
                label = labels,
                labelOptions = labelOptions(
                  style = list("font-weight" = "normal", padding = "3px 8px"),
                  textsize = "15px",
                  direction = "auto"),
                group = "Census Tracts") %>%
    addControl(title, position = "topleft") %>%
    addEasyButton(easyButton(
    icon="fa-globe", title="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }"))) %>%
    setView(lng = -122.257, lat = 47.615, zoom = 8.5)
  
  return(working_map)
  
}

```

## We are one of the "safer" Metropolitan regions
<IMG style="position:absolute; top:5%; left:80%; width:192px; "SRC="`r psrc.logo`">
```{r mpo-collisions}

mpo_chart <- metros %>% 
  mutate(injury_rate = round(injury_rate, 1)) %>%
  mutate(year = as.character(year)) %>%
  arrange(year, injury_rate) %>%
  mutate(`Traffic Related Deaths per 100k People` = injury_rate) %>%
  group_by(year) %>%
  e_charts(x=MPO_AREA, timeline=TRUE) %>%
  e_color(psrc_colors$obgnpgy_5) %>%
  e_tooltip() %>%
  e_grid(left = '15%', top = 150, bottom = 50) %>%
  e_bar(`Traffic Related Deaths per 100k People`, stack = "grp") %>%
  e_flip_coords() %>%
  e_legend(show = FALSE) %>%
  e_toolbox_feature("dataView") %>%
  e_toolbox_feature("saveAsImage") %>%
  e_title(top = 75,
          text="Five Year Average Traffic Related Deaths per 100,000 people",
          subtext=paste(sep = "\n",
                        "Collision Data Source: National Highway Traffic Safety Administration Fatality Analysis Reporting System (FARS) Data",
                        "Population Data Source: US Census Bureau American Community Survey (ACS) 5-year data Table B03002")) %>%
  e_timeline_opts(autoPlay = FALSE,
                  axis_type = "category",
                  top = 0,
                  right = 150,
                  left = 150,
                  controlStyle=FALSE,
                  lineStyle=FALSE,
                  currentIndex = 11,
                  label = list(show=TRUE,
                               color='#4C4C4C',
                               fontFamily = 'Poppins'),
                  itemStyle = list(color='#FBD6C9'),
                  checkpointStyle = list(color='#F05A28'),
                  progress = list(itemStyle = list (color='#FBD6C9')),
                  emphasis = list(itemStyle = list (color='#F05A28'))) %>%
  e_y_axis(axisLabel = list(interval = 0L)) %>%
  e_x_axis(min=0, max=16, 
           interval=2, 
           name = "Rate per 100,000 people",
           nameLocation = 'center',
           nameTextStyle = list(padding = c(10,5,5,5)))

```

<div style="position:absolute; top:15%; width: 90%;">
`r mpo_chart`
</div> 

## But Traffic Related Deaths and Serious Injuries are increasing
<IMG style="position:absolute; top:5%; left:80%; width:192px; "SRC="`r psrc.logo`">
```{r region-collisions, echo=FALSE}

region_filter <- collisions %>% 
  filter(geography=="County" & name=="Region") %>%
  select("year", "injury_type", "injuries", "injury_rate_per_100k") %>%
  pivot_longer(cols = c(injuries,injury_rate_per_100k), values_to = "estimate", names_to = "metric") %>%
  mutate(metric = str_replace_all(metric, "injuries", "Total")) %>%
  mutate(metric = str_replace_all(metric, "injury_rate_per_100k", "Rate per 100,000 people")) %>%
  pivot_wider(names_from = injury_type, values_from = estimate) %>%
  mutate(year = as.character(year))

region_plot <- 
  region_filter %>%
  group_by(metric) %>%
  e_charts(x = year, timeline=TRUE) %>%
  e_line(`Serious Injuries`, smooth = FALSE) %>%
  e_line(`Traffic Related Deaths`, smooth = FALSE) %>%
  e_color(psrc_colors$pgnobgy_5) %>%
  e_tooltip(trigger = "axis") %>%
  e_grid(left = '5%', top = 150, bottom = 50) %>%
  e_legend(show = TRUE, bottom=0) %>%
  e_toolbox_feature("dataView") %>%
  e_toolbox_feature("saveAsImage") %>%
  e_title(top = 75,
          text="Annual Traffic Related Deaths & Serious Injuries",
          subtext=paste(sep = "\n",
                        "Fatal Collision Data Source: Washington Traffic Safety Commision, Washington State Coded Fatal Crash (CFC) analytical data files",
                        "Serious Injury Data Source: Washington State Department of Transportation, Crash Data Division, Multi-Row data files (MRFF)")) %>%
  e_timeline_opts(autoPlay = FALSE,
                  axis_type = "category",
                  top = 0,
                  right = 350,
                  left = 350,
                  controlStyle=FALSE,
                  lineStyle=FALSE,
                  currentIndex = 1,
                  label = list(show=TRUE,
                               color='#4C4C4C',
                               fontFamily = 'Poppins'),
                  itemStyle = list(color='#FBD6C9'),
                  checkpointStyle = list(color='#F05A28'),
                  progress = list(itemStyle = list (color='#FBD6C9')),
                  emphasis = list(itemStyle = list (color='#F05A28'))) 

```

<div style="position:absolute; top:15%; width: 90%;">
`r region_plot`
</div> 

## Collisions are an Urban and a Rural issue
<IMG style="position:absolute; top:5%; left:80%; width:192px; "SRC="`r psrc.logo`">
```{r rural-urban}

rgeo_filter <- collisions %>%
  select("geography", "name", "year", "injury_type", "injuries") %>%
  filter(geography == "Regional Geography") %>%
  mutate(name = str_replace_all(name, "High Capacity Transit Communities", "HCT Cities")) %>%
  mutate(name = str_replace_all(name, "Urban Unincorporated", "UU")) %>%
  mutate(name = str_replace_all(name, "Metropolitan Cities", "Metro Cities")) %>%
  pivot_wider(names_from = injury_type, values_from = injuries) %>% 
  select(-"geography") %>%
  mutate(year = as.character(year), name = str_wrap(name, 7))

p1 <- 
  rgeo_filter %>%
  arrange(desc(`Traffic Related Deaths`)) %>% 
  group_by(year) %>%
  e_charts(x = name, timeline = TRUE, elementId = "chart1") %>%
  e_bar(`Traffic Related Deaths`) %>%
  e_legend(show = FALSE) %>%
  e_x_axis(axisLabel = list(interval = 0L)) %>%
  e_grid(top = 150, bottom = 50) %>%
  e_title(top = 75,
          text="Traffic Related Deaths",
          subtext=paste(sep = "\n",
                        "Source: Washington Traffic Safety Commision",
                        "Washington State CFC analytical data files")) %>%
  e_color(psrc_colors$obgnpgy_5) %>%
  e_timeline_opts(autoPlay = FALSE,
                  axis_type = "category",
                  top = 'top',
                  left = 50,
                  right= 50,
                  controlStyle=FALSE,
                  lineStyle=FALSE,
                  currentIndex = 12,
                  label = list(show=TRUE,
                               color='#4C4C4C',
                               fontFamily = 'Poppins'),
                  itemStyle = list(color='#FBD6C9'),
                  checkpointStyle = list(color='#F05A28'),
                  progress = list(itemStyle = list (color='#FBD6C9')),
                  emphasis = list(itemStyle = list (color='#F05A28'))) %>%
  e_tooltip(trigger = "axis")

p2 <- 
  rgeo_filter %>%
  arrange(desc(`Serious Injuries`)) %>% 
  group_by(year) %>%
  e_charts(x = name, timeline = TRUE, elementId = "chart2") %>%
  e_bar(`Serious Injuries`) %>% 
  e_connect(c("chart1")) %>%
  e_legend(show = FALSE) %>%
  e_x_axis(axisLabel = list(interval = 0L)) %>%
  e_grid(top = 150, bottom = 50) %>%
  e_title(top = 75,
          text="Serious Injuries",
          subtext=paste(sep = "\n",
                        "Source: Washington State Department of Transportation",
                        "Crash Data Division MRFF")) %>%
  e_color(psrc_colors$pgnobgy_5) %>%
  e_timeline_opts(show = FALSE) %>% 
  e_toolbox_feature("dataView") %>%
  e_toolbox_feature("saveAsImage") %>%
  e_tooltip(trigger = "axis")
  
rgeo_plot <- e_arrange(p1, p2, rows = 2, cols = 2) 

```

<div style="position:absolute; top:15%; width: 90%;">
`r rgeo_plot`
</div> 

## Every City is affected
<IMG style="position:absolute; top:5%; left:80%; width:192px; "SRC="`r psrc.logo`">
```{r cities}

ui <- fluidPage(
  fluidRow(
    column(12, paste0("Since 2018, ", num_cities, " of the region's 82 cities and towns (", round((num_cities/82)*100,0),"%) has had at least one serious injury or traffic related death.")),
),

br(),

  fluidRow(
    column(6, echarts4rOutput("city_wc")),
    column(6, leafletOutput("tract_collision_map"))
  )
  
)
 
server <- function(input, output){
  
  output$tract_collision_map <- renderLeaflet({create_tract_map(tracts, d_title="Fatal & Serious Injuries by Census Tract: 2018-2022")})
  
  city_filter <- reactive({
    collisions %>%
      filter(year >= 2018 & geography=="City") %>%
      select("name", "injuries") %>%
      group_by(name) %>%
      summarise(freq=sum(injuries)) %>%
      as_tibble
  })
  
  output$city_wc <- renderEcharts4r({
    city_filter() %>%
      e_charts() %>%
      e_cloud(name, freq, shape = "rectangle", sizeRange = c(1, 75)) %>%
      e_toolbox_feature("dataView") %>%
      e_toolbox_feature("saveAsImage") %>%
      e_title(text=paste(sep = "\n","Cities with Traffic Related Deaths",
                         "or Serious Injuries"),
              padding = list(c(20,5,20,5)))

  })
  
}

shinyApp(ui, server)

```


## Traffic Deaths Disproportionally Impact People of Color

```{r racial-disparities}

ui <- fluidPage(
  fluidRow(
    column(3, ""),
    column(6,
           sliderInput("race_year", label = NULL, sep="", animate=FALSE, 
                       min = min(collisions$year), max = max(collisions$year), 
                       value = 2021)),
    column(3, "")),
  
  fluidRow(
    column(12, echarts4rOutput("racial_disparities_plot"))
  ),
  
    fluidRow(
    column(12, "Source: Washington Traffic Safety Commision, Washington State Coded Fatal Crash (CFC) analytical data files")
  )
  
)
 
server <- function(input, output){
  
  fatal_filter <- reactive({
    collisions %>%
      select("geography", "name", "year", "injury_type", fatality_rate="injury_rate_per_100k") %>%
      filter(geography == "Race & Ethnicity" & injury_type == "Traffic Related Deaths" & year==input$race_year) %>%
      arrange(fatality_rate) %>%
      mutate(race= str_wrap(name, 15))
  })
  
  output$racial_disparities_plot <- renderEcharts4r({
    fatal_filter() %>% 
    e_charts(x=race) %>%
    e_color(psrc_colors$pgnobgy_5) %>%
    e_tooltip() %>%
    e_grid(left = '20%') %>%
    e_pictorial_(serie = "fatality_rate", 
                symbol = fa_user,
                symbolRepeat = TRUE, z = -1,
                symbolSize = 30,
                legend = FALSE,
                name="Fatality Rate per 100,000 people") %>%
    e_flip_coords() %>%
    e_legend(show = FALSE) %>%
    e_x_axis(splitLine=list(show = FALSE)) %>% 
    e_y_axis(splitLine=list(show = FALSE),
             axisPointer = list(show = FALSE)) %>%
    e_toolbox_feature("dataView") %>%
    e_toolbox_feature("saveAsImage") %>%
    e_title("Traffic Related Deaths")

  })
}

shinyApp(ui, server)

```

## Men acccount for a large share of injuries and deaths

```{r gender-disparities}

ui <- fluidPage(
  fluidRow(
    column(3, ""),
    column(6,
           sliderInput("gender_year", label = NULL, sep="", animate=FALSE, 
                       min = min(collisions$year), max = max(collisions$year), 
                       value = max(collisions$year))),
    column(3, "")),
  
  fluidRow(
    column(6, echarts4rOutput("gender_serious_plot")),
    column(6, echarts4rOutput("gender_fatal_plot"))
  ),
  
    fluidRow(
    column(12, "Source: Washington Traffic Safety Commision, Washington State Coded Fatal Crash (CFC) analytical data files")
  )
  
)
 
server <- function(input, output){
  
  gender_fatal_filter <- reactive({
    collisions %>% 
      select("geography", "name", "year", "injury_type", "injuries") %>%
      filter(geography == "Gender" & injury_type == "Traffic Related Deaths" & year==input$gender_year)
  })
  
  gender_serious_filter <- reactive({
    collisions %>% 
      select("geography", "name", "year", "injury_type", "injuries") %>%
      filter(geography == "Gender" & injury_type == "Serious Injuries" & year==input$gender_year)
  })
  
  output$gender_fatal_plot <- renderEcharts4r({
    gender_fatal_filter() %>% 
      rename(`Traffic Related Deaths` = "injuries") %>%
      e_charts(name) %>%
      e_pie(`Traffic Related Deaths`, radius = c("50%", "70%")) %>% 
      e_color(psrc_colors$pgnobgy_5) %>%
      e_toolbox_feature("dataView") %>%
      e_toolbox_feature("saveAsImage") %>%
      e_tooltip() %>%
      e_title("Traffic Related Deaths") %>%
      e_legend(show = FALSE)

  })
  
  output$gender_serious_plot <- renderEcharts4r({
    gender_serious_filter() %>% 
      rename(`Serious Injuries` = "injuries") %>%
      e_charts(name) %>%
      e_pie(`Serious Injuries`, radius = c("50%", "70%")) %>% 
      e_color(psrc_colors$pgnobgy_5) %>%
      e_toolbox_feature("dataZoom") %>%
      e_toolbox_feature(feature="reset") %>%
      e_toolbox_feature("dataView") %>%
      e_toolbox_feature("saveAsImage") %>%
      e_tooltip() %>%
      e_title("Serious Injuries") %>%
      e_legend(show = FALSE)

  })
  
  
}

shinyApp(ui, server)

```

## Younger People are more likely to be in a Serious Collisions

```{r age-disparities}

ui <- fluidPage(
  fluidRow(
    column(3, ""),
    column(6,
           sliderInput("age_year", label = NULL, sep="", animate=FALSE, 
                       min = min(collisions$year), max = 2021, 
                       value = 2021)),
    column(3, "")),
  
  fluidRow(
    column(6, echarts4rOutput("age_total_plot")),
    column(6, echarts4rOutput("age_rate_plot"))
  ),
  
    fluidRow(
    column(12, "Source: Washington Traffic Safety Commision, Washington State Coded Fatal Crash (CFC) analytical data files")
  )
  
)
 
server <- function(input, output){
  
  age_total_filter <- reactive({
    collisions %>% 
      filter(geography=="Age" & year==input$age_year & name != "Unknown") %>%
      select("name", "injury_type", value="injuries")
  })
  
  age_rate_filter <- reactive({
    collisions %>% 
      filter(geography=="Age" & year==input$age_year & name != "Unknown") %>%
      select("name", "injury_type", value="injury_rate_per_100k")
  })
  
  output$age_total_plot <- renderEcharts4r({
    age_total_filter() %>% 
      group_by(injury_type) %>%
      e_charts(name) %>%
      e_bar(value, stack = "grp") %>%
      e_color(psrc_colors$pgnobgy_5) %>%
      e_toolbox_feature("dataZoom") %>%
      e_toolbox_feature(feature="reset") %>%
      e_toolbox_feature("dataView") %>%
      e_toolbox_feature("saveAsImage") %>%
      e_tooltip(trigger = "axis") %>%
      e_x_axis(axisTick=list(show = FALSE)) %>%
      e_show_loading() %>%
      e_title("Total") %>%
      e_legend(show = FALSE)
  })
  
  output$age_rate_plot <- renderEcharts4r({
    age_rate_filter() %>% 
      group_by(injury_type) %>%
      e_charts(name) %>%
      e_bar(value, stack = "grp") %>%
      e_color(psrc_colors$pgnobgy_5) %>%
      e_toolbox_feature("dataView") %>%
      e_toolbox_feature("saveAsImage") %>%
      e_tooltip(trigger = "axis") %>%
      e_x_axis(axisTick=list(show = FALSE)) %>%
      e_show_loading() %>%
      e_title("Rate per 100,000 People") %>%
      e_legend(show = FALSE)
  })
  
  
}

shinyApp(ui, server)

```

